---
title: "Homework 2"
output: html_document
author: "Amanda Jimenez"
---

RULES:
1. Deadline for homework is Sunday 23:59. Try to turn it in at least a few minutes before to avoid any issues with failed uploads, slow internet, etc. Late admissions are not accepted.
2. Backup your work, preferrably using GitHub.
3. If you get stuck on something, try googling it first. You can find answers to a lot of questions on several public forums, but consider those to be guidelines and not solutions. ABSOLUTELY DO NOT COPY-PASTE THE ANSWERS FROM THE INTERNET.
4. If that fails, you're welcome to ask a question on the Moodle forum, but try to do it in a way which doesn't explain or hint at a part of a solution. If you cannot ask your question without revealing a part of the solution, send an e-mail to dglavas@bioinfo.hr instead of posting on the forum. 
5. ABSOLUTELY DO NOT COPY-PASTE THE ANSWERS FROM YOUR COLLEAGUES.
6. Each task states how the final solution (variable or function) should be named. Please take care to name the answers exactly as specified because it helps us in correcting your homeworks.
6. Please hand in a "clean" version of your homework. That means no error-producing code and no more than one solution for each task (unless specified otherwise in the question). Also, please leave out or at least comment parts of the code which are specific to your system (such as setting the work directory and installing packages). When you're ready to hand in the final version of your homework, please run it once more in a clean environment to make sure you have all variables defined. This means cleaning your environment (either by hand or by re-starting your R session) and then running the whole code once in the order in which it's written.
7. Do not use any additional packages besides what is in base R (so already loaded automatically when you start a new R session) unless given explicit permission to do so.


  
### 1.
##### 5 points
Create a binary operator approx (modify the name if you need to) which will compare two numeric vectors. The operator should compare numbers in the same way as ==, but with tolerance of +-0.1 (meaning TRUE if the difference in value is 0.1 or smaller, FALSE otherwise).
```{r}
`%approx%` <- function(x,y){
  dif <- x-y
  equals <- ifelse(dif<=0.1, 'TRUE', 'FALSE')
  equals 
}
x <- c(1.01, 1.1, 1.3)
y <- c(1.01, 1.05, 1)
x %approx% y
```
<font color="red">You don't need ifelse() here; just " equals <- dif<=0.1 " would do the same thing. This is not a mistake, just stylistic advice for code readability. But outputting TRUE and FALSE as character strings and not boolean variables is a mistake - you can't directly subset anything using the output of your function! Never treat TRUE and FALSE as characters without a very very good reason. 3/5 </font>



### 2.  
##### 18 points
In this task, you will continue to work on the examples from the second lecture. You can use or further modify any function we created during this lecture (any, so not just the final ones!).

A) Different coin dimensions are as follows: radii from 1.0 to 2.2 cm in increments of 0.1, thickness from 0.1 to 0.3 in increments of 0.02. Make every possible combination of those elements - so all possible pairs of radius and thickness. Save this as a single variable. Mark clearly what is radius and what is thickness.
Hint: there is a built-in R function which produces all unique combinations. It's in base R, so you shouldn't need to install any new packages. (2 points)
```{r}
radius <- seq(1.0, 2.2, by = 0.1)
thickness <- seq(0.1, 0.3, by = 0.02)
dimensions <- list(radius = radius, thickness = thickness)
grid <- expand.grid(dimensions)
grid
```

B) Next, create vectors of coin counts for each combination - this represents count per each type of coin per square (number of coins of one type found in one square is always between 50 and 300, chosen at random). Make the vectors differ in length because not every type of coin is found in each square. Mean vector length should be around 500, with standard deviation approximately 100. Note the "around" and "approximately" - this is not exactly normal distribution because all lengths are integers, but it should approximate normal distribution. Figure out how to do this, do not use uniform distribution!   
Save this information together with dimension combinations in one object (variable). This object can be of any type you find to be appropriate, but bear in mind that 1. you will have to work with it later and 2. it has to be clear which vector "belongs" to which radius-thickness combination. (3 points)  
```{r}
set.seed(143)
coin_count <- replicate(143, list(toString(as.integer(runif(round(rnorm(1, 500, 100), 0), min = 50, max = 300)))))
grid$coin.count <- as.character(coin_count)
grid
```

<font color="red">Why save the counts as characters? You could've left them as numeric vectors, add them to grid in the same way as you did with character vectors/strings, and save yourself the trouble of string splitting and conversion later. Conversions are super-convenient when you need them, but don't do them if they're not necessary. Elements in one column of data frame don't have to have length 1, they can be vectors as well (as we said, data frames are basically lists internally so you can save pretty much anything, as long as the number of elements/rows matches across the table).  2.5/3
  
Also, a note regarding sampling: if you need 143 vector lengths, it's better to pull them from desired distribution at once. Functions rnorm, runif, and similar guarantee to pull a sample from a particular distribution, but not one of them "remembers" the last iteration when they were called. In other words, it is never a good idea to split your sampling into multiple iterations - because the bigger the sample you're pulling at once, the surer you can be that it really fits the desired distribution. Random number generators are far from perfect, computationally speaking, so try to stick to bigger samples.  </font>


C) Make a function which will calculate total weight of gold coins of various dimensions. The function's input should be an object of the same type and layout as the object you created in part B) - in other words, your new function should work on the object you created. The function should return one number: total weight. (4 points)
```{r}
coinWeight <- function(df) {
  d_gold <- 19.3
  r <- as.vector(df[,1])    ### why do you need as.vector() here? it's already a vector
  th <- as.vector(df[,2])
  n <- sapply(strsplit(as.character(df[, 3]), ','), function(x) sum(as.numeric(x)))
  weight <- mapply(function(r,th,n){vol_one <- (r^2) * pi * th; vol <- sum(vol_one * n); weight <- vol * d_gold / 1000; return(weight)}, r, th, n)
    total_weight <- sum(weight)
    total_weight
}
coinWeight(grid)
```
<font color="red">nice usage of mapply!</font>

D) Generate data for gold, silver and copper coins. Silver coins have radii from 1.5 to 3.0 in increments of 0.25 and thickness from 0.12 to 0.28 in increments of 0.02. Copper coins have radii from 1.0 to 2.2 cm in increments of 0.2 and thickness from 0.15 to 0.35 in increments of 0.05. You can assume per-square and total distributions are the same as for gold coins in B).
Basically, you need to repeat the data creation process from A) and B) three times with different parameters. Do NOT do this by writing almost the same code three times! You need to find a way to "automate" the process. You can choose to organise everything from A) and B) into one function, or split the the process into smaller parts and handle it that way - whatever you find easiest. But you need to generate all the data with one call of (each) function. You will probably use some kind of repetition or looping mechanism, but use only the ones which are optimised in R (e.g. apply family). You are not allowed to use classic for/while/repeat loops. (5 points)
```{r}
Dimensions <- function(r, th){
  dimensions <- list(radius = r, thickness = th)
  grid <- expand.grid(dimensions)
  return(grid)
}
Generate_vectors <- function(grid){
  rows <- nrow(grid)
  set.seed(100)
  coin_count <- replicate(rows, list(toString(as.integer(runif(round(rnorm(1, 500, 100), 0), min = 50, max = 300)))))
  grid$coin.count <- as.character(coin_count)
  return(grid)  
}
Generate_data <- function(material){
  if(material == "gold"){
    r <- seq(1.0, 2.2, by = 0.1)
    th <- seq(0.1, 0.3, by = 0.02)
  }
  if(material == "silver"){
    r <- seq(1.5, 3.0, by = 0.25)
    th <- seq(0.12, 0.28, by = 0.02) 
  }
  if(material == "copper"){
    r <- seq(1.0, 2.2, by = 0.2)
    th <- seq(0.15, 0.35, by = 0.05)
  }
grid <- Dimensions(r, th)                ### take notice of indentation - this line and the next one should be indented so it's 
df_material <- Generate_vectors(grid)    ### clearly visible they're a part of function body 
}
materials <- c("gold", "silver", "copper")
three_df <- lapply(materials, Generate_data)
all_types_df <- Reduce(function(x, y) merge(x, y, all=TRUE), lapply(materials, Generate_data), accumulate=FALSE)
three_df
```

<font color="red">Very nice function organisation and later usage of lappy(), but merge is not the right chioce in the last step. It worked because you specified all=TRUE and there is no combination of dimensions and counts exactly the same so it just bound all three data frames together row by row. However, if you by chance had the same row appearing in 2 or more data frames, you would've lost it. merge() is designed to join tables by reference columns - intersecting them in a way. What you want here is to flatten the list and get 3 dataframes "pasted" together into one (and preferably not lose the information on which row pertains to which material, which you did). In this case, Reduce(rbind, three_df)  would've been the way to go. Reduce-merge is brilliant if you want to sequentially merge several data frames together (in other words, add to one table some columns from another, but not by position but in the way that common information still matches), but unnecessarily risky in this case. Other than that, perfect solution. 4/5  </font>


E) Apply the function you created in C)* on the whole data set created in D). Calculate total weights for each type of metal separately, and the total weight you need to haul from the dragon's cave (all the coins together). You should return all that information in one object. (4 points)
*You will probably need to adjust your original function to include additional parameters. You don't need to re-write the whole function here if you don't want to, you can go back to C) and change the necessary parts there - but remember to check if it still functions correctly!
```{r}
coinWeight <- function(df, x) {
  d_gold <- 19.3
  d_silver <- 10.5
  d_copper <- 9.0
  r <- as.vector(df[,1])
  th <- as.vector(df[,2])
  n <- sapply(strsplit(as.character(df[, 3]), ','), function(x) sum(as.numeric(x)))
  if(x==1){
    weight <- mapply(function(r,th,n){vol_one <- (r^2) * pi * th; vol <- sum(vol_one * n); weight <- vol * d_gold / 1000; return(weight)}, r, th, n)
  }
  if(x==2){
    weight <- mapply(function(r,th,n){vol_one <- (r^2) * pi * th; vol <- sum(vol_one * n); weight <- vol * d_silver / 1000; return(weight)}, r, th, n)
  }
  if(x==3){
    weight <- mapply(function(r,th,n){vol_one <- (r^2) * pi * th; vol <- sum(vol_one * n); weight <- vol * d_copper / 1000; return(weight)}, r, th, n)
  }
  total_weight <- sum(weight)
  total_weight
}
positions <- which(lengths(three_df) > 0)   ### ??? this is unnecessarily complicated, especially considering the fact that you don't actually need this type of counter
weights <- c(mapply(coinWeight,three_df, positions), sum(weight_total))
types <- c("gold_coins_weight", "silver_coins_weight", "copper_coins_weight", "total_coins_weight")
final_weights <- data.frame(types, weight_total)
final_weights
```

<font color="red">The three ifs (and the accompanying code repetition) could've been avoided if you defined density as an input. You're already using mapply and can pass multiple inputs, so why not take advantage of that?  
Unfortunately, variable weight_total doesn't exist so I can't run this chunk. I'm guessing that at one point mapply(coinWeight,three_df, positions) was saved as weight_total, and then the sum was added at the end of that vector (so I'll do it like that to be able to check), but that code is gone now and you didn't check if everything runs in a clean environment.  3/4 because everything else is correct.  
Also, a note: named vector would be a better choice for returning weights information. Same info, less complicated to handle later. </font>


### 3.  
#### 4 points
Create two functions, getSum1 and getSum2, both of which will take 2 arguments (numbers m and n) and calculate the sum of integer numbers from m to n. For example, the result of calling the function with parameters 4 and 6 should be 15 (4 + 5 + 6 = 15). You should get the same result if you called the function with 6 and 4 (in different order). Solve in two different ways.
```{r}
getSum1 <- function(m, n){
  vector <- seq(m, n, length=abs(m-n)+1)
  result <- sum(vector)
  result
}
getSum1(6,4)

getSum2 <- function(m,n){
  ifelse(m<n, result <- (m+n)/2 * (n-m+1), result <- (n+m)/2 * (m-n+1))
  result
}
getSum2(4,6)
```
<font color="red">nice!</font>



### 4.  
#### 5 points
You are an engineer in the Star Fleet, tasked with re-calibrating the anti-collision system. You decide to approximate spaceships as vectors in 3D coordinate space.
(You are not allowed to use function dist() in this task. Write formulas for distance calculation yourself!)

A) Create a function manhattan that calculates Manhattan distance between two vectors.
```{r}
manhattan <- function(x, y){
     dist <- abs(x-y)
     dist <- sum(dist)
     dist
}
x <- c(1, 7, 2)
y <- c(2, 3, 3)
manhattan(x,y)
```

B) Create a function euclidean that calculates Euclidean distance between two vectors.
```{r}
euclidean<-function(x, y){
  return(sqrt(sum(x-y)^2))
}
x <- c(1, 7, 2)
y <- c(2, 3, 3)
euclidean(x,y)
```

C) Create a function cheby that calculates Chebyshev distance between two vectors.
```{r}
cheby <- function(x, y) {
  max(abs(x - y))
} 
x <- c(1, 7, 2)
y <- c(2, 3, 3)
cheby(x,y)
```

D) Create a function getDistance with three parameters (v1, v2, metric) that will calculate a distance between vectors v1 and v2. The distance to be used will be passed as a string via metric argument and can be one of the following: “Manhattan”, “Euclidean”, “Chebyshev”. Use the functions you created before inside the final function. Add a warning if distance is smaller than 3.
```{r}
getDistance <- function(v1, v2, metric){
  if(metric=="Manhattan"){
    result<-manhattan(v1, v2)
  }
  if(metric=="Euclidean"){
    result<-euclidean(v1, v2)
  }
  if(metric=="Chebyshev"){
    result<-cheby(v1, v2)
  }
  else{
    return("Argument 'metric' not recognisible")
  }
  result
}
v1 <- c(1, 7, 2)
v2 <- c(2, 3, 3)
getDistance(v1, v2, "Chevyshev")
```

<font color="red">Everything good except distance warning is missing. 4/5   </font>


### 5.  
#### 3+2 points
A) Make a function est_pi that will estimate the value of the number pi by randomly generating points inside a square and checking whether they fall within a circle inscribed in that square. Points fall within a circle if the distance from circle center is smaller than or equal to the radius of the circle.  
```{r}
est_pi <- function(N, R){ #N is the number of points we want, the number of estimations we will make. R is the ratio of the circle inscribed in a square with the 2*R side length.
  x <- runif(N, min= -R, max= R)
  y <- runif(N, min= -R, max= R)
  in_or_out <- (x^2 + y^2) <= R^2 #returns TRUE if the point is inside one quarter of the circle (<=R), and FALSE if the point lands inside the square with 2*R side length but outside the circle (>R)
  estimate <- 4 * sum(in_or_out) / N #sum of all the TRUE values (which value are equivalent to 1), *4 to have the whole circle, and /N to have the actual probability of each point.
  estimate
}
est_pi(10000, 4)
```

B) Repeat this estimation 100,000 times using function replicate (if the computer on which you're working on can't handle it, lower the number of repetitions; if this function call runs longer than 5-10 minutes, stop it, lower the numbers, and try again). How much (in % terms) is it different from actual value of pi if you first set the seed to 1234? (Google function set.seed() and what it does.) Use the pi variable that is part of the base R package to do this comparison. 
```{r}
set.seed(1234)
repetition <- replicate(100000, est_pi(10000, 4))
pi_mean <- mean(repetition)
dif_percentage <- abs((pi_mean-pi)/pi)*100
dif_percentage
# In my case, a value of difference percentage quite low (0.00115%), so it is a good estimation.
```

<font color="red">Yes, it's a good estimation, and well-written as well. Great! </font>

<font color="red">IMPORTANT NOTICE  
Few of your functions have a striking resemblance to the code one of your colleagues submitted. You obviously invested more time in your homework so I can speculate about the direction in which the copy-paste went, but it doesn't matter - next time we see something like that, everybody involved loses all points for the task, and gets additional negative points as well. This is your first and final warning. Please take it seriously.
</font>
